---
title: "Reproducible Research: Peer Assessment 1"
author: "Devender R. Gollapally"
output: 
  html_document:
    keep_md: true
    
---

## Introduction

Using the data from a personal activity monitoring device we will try to answer the following questions:

1. **What is mean total number of steps taken per day ?**
2. **What is the average daily activity pattern ?**
3. **Are there differences in activity patterns between weekdays and weekends ?**

The device used, collects data at 5 minute intervals. The data consists of two months collected during the months of October and November, 2012 and include the number of steps taken each day.

## Data Source

The data used in this assignment can be downloaded from this [Link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) on the course website, it is also included in the [github repository](https://github.com/devender/RepData_PeerAssessment1).

##### R workspace setup
```{r warning=FALSE,message=FALSE,results="hide"}
library(data.table)
library(dplyr)
library(ggplot2)
library(lubridate)

# Setup Working dir
setwd("/Users/Devender/RepData_PeerAssessment1")
# Clean up work space
rm(list=ls())
gc()
if(!file.exists("activity.csv")){
    unzip("activity.zip")
}
```

## Loading and preprocessing the data
Reading the CSV file into a data frame so that we can use if for later processing.
```{r}
activityData <- read.csv("activity.csv", header= T, na.strings = c("NA"), stringsAsFactors = F)
str(activityData)
```

## What is mean total number of steps taken per day ?

Subsetting the data and grouping by each day, gives us total number of steps for each day.

```{r warning=FALSE}
stepsByDay <- activityData %>% 
                select(steps, date) %>% 
                filter(!is.na(steps)) %>%
                group_by(date) %>% 
                summarise(total_steps = sum(steps))
head(stepsByDay)
```
Using the above data we can plot a **histogram** of the total number of steps taken each day

```{r warning=FALSE}
ggplot(stepsByDay, aes(x=total_steps)) + 
    geom_histogram(colour="black", fill="white", bins=10)+
    xlab("Steps") +ylab("Frequency") +
    ggtitle("Total number of steps taken each day")
```

##### Calculate and report the mean and median of the total number of steps taken per day:
```{r}
mean(stepsByDay$total_steps, na.rm =TRUE)
median(stepsByDay$total_steps, na.rm=TRUE)
```


## What is the average daily activity pattern?
Lets compute during what times of the day the person is more active.
```{r}
avgActivityPattern <- activityData %>%
    filter(!is.na(steps)) %>%
    group_by(interval) %>% 
    summarise(avg_steps = mean(steps))

ggplot(avgActivityPattern, aes(x=interval)) + 
    geom_line( aes(y=avg_steps)) +
    xlab("Interval of Day") +ylab("Average Number of Steps") +
    ggtitle("Avg Daily Activity Pattern")
```

##### The 5-minute interval that, on average, contains the maximum number of steps:
```{r}
avgActivityPattern %>% arrange(desc(avg_steps)) %>% select(interval) %>% head(1) %>% as.numeric
```
## Imputing missing values

Lets find out how just many rows have NA.

##### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with 𝙽𝙰s):
```{r}
numOfRowsWithNA <- activityData %>% 
                filter(is.na(steps)) %>% 
                summarise(n()) %>% 
                as.numeric()

totalRows <- nrow(activityData)
numOfRowsWithNA
```
The Number of rows with NA is `r numOfRowsWithNA` it represents `r (numOfRowsWithNA/totalRows)*100` percent. Since this is a rather significant percentage we will fill in this missing data with out best guess.

##### Strategy for filling in all of the missing values in the dataset:

Create a new dataset with fixed na steps replaced with 0.

```{r}
activityDataFixed = within(activityData, {
    steps = ifelse(is.na(steps), 0 , steps)
})
str(activityDataFixed)
```
Now that we have created a new column with the mising steps fixed lets recreate the histogram to see if there is a difference

```{r warning=FALSE}
stepsByDayFixed <- activityDataFixed %>% 
                select(steps, date) %>% 
                group_by(date) %>% 
                summarise(total_steps = sum(steps))
head(stepsByDayFixed)
```
Using the above data we can plot a **histogram** of the total number of steps taken each day

```{r warning=FALSE}
ggplot(stepsByDayFixed, aes(x=total_steps)) + 
    geom_histogram(colour="black", fill="white", bins=10)+
    xlab("Steps") +ylab("Frequency") +
    ggtitle("Total number of steps taken each day")
```

##### Calculate and report the mean and median of the total number of steps taken per day:
```{r}
mean(stepsByDayFixed$total_steps)
median(stepsByDayFixed$total_steps)
```

### Impact of imputing missing data

We see that by imputing the missing data with 0 effects both the density functions (mean and median).

## Are there differences in activity patterns between weekdays and weekends?
First we will add another column to indicate if it is a weekend or a weekday.

```{r}
isWeekend <- function(dt){
    if(weekdays(ymd(dt)) %in% c("Sunday","Saturday")){
        return("WEEKEND")
    }else{
        return("WEEKDAY")
    }
}

activityDataFixed=activityDataFixed %>% rowwise() %>% mutate(wday = isWeekend(date))
```
Next we will plot Average by weekday and weekend.

```{r}
compareWeekdayVsWeekend <- activityDataFixed %>% 
    group_by(wday,interval) %>% 
    summarise(mean=mean(steps))

ggplot(compareWeekdayVsWeekend, aes(x = interval, y = mean)) + ylab("Number of Steps") + geom_line() + facet_grid(wday~.)
```